---
title: "Initial Exprolatory Data Analysis"
author:
- name: Elyas Heidari & Artem Lomakin
  affiliation: 
  - AI in Oncology, DKFZ Heidelberg
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
    number_sections: false
---

# Setup / definitions

## Libraries

```{r setup_knitr, include=FALSE}
library('BiocStyle')
options(bitmapType='cairo')
set.seed(1996)
knitr::opts_chunk$set( autodep=TRUE, cache=TRUE, cache.lazy=FALSE, dev='png' )
knitr::opts_knit$set( root.dir='..' )
# wflow_build(files='analysis/00_initial_eda.Rmd', view=F, verbose=T, delete_cache=T)
```

```{r setup_libs, collapse=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
suppressMessages(source('code/R/dep.R')) 
```

## Helper functions

```{r setup_helpers, message=FALSE, cache=TRUE}
suppressMessages(source('code/R/utils.R')) 
library("spatialLIBD")
# source('code/R/sma01_eda.R')
```

## Paths

```{r setup_input}
tag       = '00_initial_eda' 
# ref_tag   = 'embryo1_2'
data_raw_dir  = 'data_raw'
data_tidy_dir = 'data_tidy'
out_dir   =  file.path('output', tag)

decodes_dir = 'data_raw/decodes'
decodes_f   = list.files(decodes_dir, full.names=TRUE)
decodes_tags = decodes_f %>% 
    purrr::map(~str_replace(.x, pattern='data_raw/decodes/', replacement='')) %>% 
    purrr::map(~str_replace(.x, pattern='_DECODEv1.csv', replacement='')) %>% 
    unlist
names(decodes_f) = decodes_tags

nodules_anno_dir = 'data_raw/nodule_annotations'
nodules_anno_f   = list.files(nodules_anno_dir, full.names=TRUE)
nodules_anno_tags = nodules_anno_f %>% 
    purrr::map(~str_replace(.x, pattern='data_raw/nodule_annotations/', replacement='')) %>%
    purrr::map(~str_replace(.x, pattern='\\.csv', replacement='')) %>%
    unlist
names(nodules_anno_f) = nodules_anno_tags


samples_f     = list.dirs(data_raw_dir, recursive=FALSE) %>% str_subset('ML')
samples_tags  = samples_f %>% purrr::map(~str_replace(.x, pattern='data_raw/', replacement='')) %>% unlist
names(samples_f) = samples_tags

obj_dir  = file.path('objects', tag)
dir.create(obj_dir)
obj_list = list()
fig_dir  = file.path('output', tag)
dir.create(fig_dir)
fig_list = list()
```


## Load inputs

```{r load_inputs, fig.height=8, fig.width=8, results='asis', cache=TRUE}
remove_empty_spots <- function(spe){
    spe = spe[,colSums(counts(spe)) != 0]
    spe
}

remove_isolated_nodes <- function(ig){
    Isolated = which(igraph::degree(ig)==0)
    delete.vertices(ig, Isolated)
}
samples = samples_f %>% 
    purrr::map(read10xVisium) %>%
    purrr::map(remove_empty_spots) %>%
    purrr::map(scuttle::logNormCounts)

samples_raw = samples_f %>% 
    purrr::map(read10xVisium) %>%
    purrr::map(remove_empty_spots) %>%
    purrr::map(scuttle::logNormCounts)

hvgs = samples %>%
    map(modelGeneVar) %>%
    map(getTopHVGs) %>%
    map(~ .x[1:1000]) %>%
    purrr::reduce(intersect)


decodes = decodes_f %>%
    map(fread) 

markers = c('Hnf4a', 'Krt19', 'Lrat', 'Eng', 'Ptprc', 'Adgre1', 'Csf1r', 'Ccr2', 'Itgax', 'Cd209a', 'Itgae', 'Siglech', 'Cd40', 'Cd86', 'Ngp', 'Ncr1', 'Gzma', 'Nkg7', 'Cd3d', 'Cd8a', 'Tcf7', 'Lef1', 'Gzmk', 'Mki67', 'Cd4', 'Tcrg-C1', 'Cd79a', 'Ms4a1')

gene_data = rowData(samples[[1]])
gene_data$ens_id = rownames(gene_data)
rownames(gene_data) = gene_data$symbol
marker_ens = gene_data[markers,]$ens_id

sel_genes = c(hvgs, decodes[['Plasmid']]$ID, marker_ens) %>% unique  

samples %<>% purrr::map(~.x[sel_genes,])

nodules_anno = nodules_anno_f %>% 
    map(fread) %>%
    map(~setnames(.x, 2, 'nodule')) %>%
    map(~.x[, nodule := ifelse(nodule != '', nodule, 'unassigned')])

exprs_list = samples %>% 
    # map(~.x[sel_genes,]) %>%
    map(logcounts) %>%
    map(as.matrix) %>%
    map(~.x[rowSums(.x) != 0,]) 

for(sample_tag in samples_tags){
    rownames(exprs_list[[sample_tag]]) = rowData(samples[[sample_tag]])[rownames(exprs_list[[sample_tag]]),]
}

for(sample_tag in samples_tags){
    sample_coldt = colData(samples[[sample_tag]]) %>% as.data.table
    sample_coldt$Barcode = colnames(samples[[sample_tag]])
    sample_coldt %<>% .[nodules_anno[[sample_tag]], on='Barcode']
    sample_coldf = DataFrame(sample_coldt)
    rownames(sample_coldf) = sample_coldt$Barcode
    sample_coldf %<>% .[colnames(samples[[sample_tag]]),]
    colData(samples[[sample_tag]]) = sample_coldf
}

shared_genes = exprs_list %>% map(rownames) %>% purrr::reduce(intersect)
cons_adj = exprs_list %>%
    map(~.x[shared_genes,]) %>%
    map(t) %>%
    map(data.table) %>%
    map(~cov.wt(.x, method='ML')$cov) %>%
    map(cov2cor) %>%
    map(glasso::glasso, rho=0.1) %>%
    map(~.x$wi) %>%
    purrr::reduce(`+`) %>%
    `/`(length(exprs_list))
cons_adj_bin =  abs(cons_adj) > 0.075
diag(cons_adj_bin ) = F
rownames(cons_adj_bin ) = colnames(cons_adj_bin) = shared_genes
cons_graph = graph_from_adjacency_matrix(cons_adj_bin) %>%
    as.undirected %>%
    remove_isolated_nodes
# cons_graph = as.undirected(cons_graph)

cons_mem = cons_graph %>%
    cluster_louvain %>%
    membership




cons_lay = layout_nicely(cons_graph)
graph_col_comm(
    graph  = cons_graph, 
    lay    = cons_lay, 
    grp    = .palette1[cons_mem], 
    title  = 'Consensus graph on all slides', 
    labels = names(cons_mem)
)

add_genes = setdiff(markers, cons_mem)
add_mem   = rep(max(cons_mem)+1, length(add_genes))
names(add_mem) = add_genes
cons_mem  = c(cons_mem, add_mem)

gg = exprs_list %>%
    map(t) %>%
    map(get_network, rho = .1, threshold = .075) %>%
    map(as.undirected) %>%
    map(remove_isolated_nodes)


memberships = gg %>%
    map(cluster_louvain) %>%
    map(membership)

act_all_nodules  = samples_tags %>% 
    map(~act_colors(t(exprs_list[[.x]]), samples[[.x]]$nodule))
names(act_all_nodules) = samples_tags



adata_q = zellkonverter::readH5AD('data_tidy/integrated_query_nodules_umapped.h5ad')

col_dt_all = colData(adata_q) %>%
    as.data.table %>% setkey('Barcode') %>% 
    .[,c('Barcode', 'section', 'leiden_sagenet')]


for(sample_tag in samples_tags){

       print(sample_tag)
       col_dt_1 = colData(samples[[sample_tag]]) %>% as.data.table %>% setkey('Barcode')
       col_dt_2 = col_dt_all[section==sample_tag]
        colData(samples[[sample_tag]]) = 
            col_dt_2[col_dt_1, all=TRUE] %>%
            .[, leiden_sagenet := ifelse(is.na(leiden_sagenet), 'normal', leiden_sagenet)] %>% 
            DataFrame
}

act_all_leiden  = samples_tags %>% 
    map(~act_colors(t(exprs_list[[.x]][names(cons_mem),]), samples[[.x]]$leiden_sagenet))
names(act_all_leiden) = samples_tags

# act_all_leiden  = samples_tags %>% 
#     map(~act_colors(t(exprs_list[[.x]]), samples[[.x]]$leiden_sagenet))
# names(act_all_leiden) = samples_tags

# lay = layout_nicely(gg[[1]])

# for(nodule in rownames(act_all_nodules)){
#     graph_col_act(gg[[1]], act_all_nodules[nodule, names(V(gg[[1]]))], lay, sprintf('%s: %s', samples_tags[1], nodule))
# }

# graph_col_comm(
#     graph  = gg[[1]], 
#     lay    = lay, 
#     grp    = .palette1[memberships[[1]]], 
#     title  = 'GGM on ML_I', 
#     labels = names(memberships[[1]])
# )



# samples_tags %>%
#     map(~SingleCellExperiment(
#         assays = list(counts=exprs_list[[.x]][shared_genes,]),
#         colData = colData(samples[[.x]]),
#         # rowData = rowData(samples[[.x]]),
#         metadata = list(adj = cons_adj_bin * 1, adj_local=as_adjacency_matrix(gg[[.x]]))
#     ) %>%
#     writeH5AD(file = sprintf(file.path(data_tidy_dir, '%s/%s.h5ad'), tag, .x))
# )


# SingleCellExperiment(
#         assays = list(counts=exprs_list[[1]][shared_genes,]),
#         colData = colData(samples[[1]]),
#         # rowData = rowData(samples[[1]]),
#         metadata = list(adj = cons_adj_bin * 1, adj_local=as_adjacency_matrix(gg[[1]]))
#     ) %>% writeH5AD('data_tidy/test.h5ad')
# data_tidy/01_sagenet_integration/integrated_query_umapped.h5ad

# for(mem in unique(cons_mem)){
#     gene_list = names(which(cons_mem == mem))
#     res = enrichr(genes = gene_list, databases = "PanglaoDB_Augmented_2021")[[1]]

#     par(mfrow = c(1, 1), mar = c(3, 25, 2, 1))
#     barplot(height = -log10(res$P.value)[5:1], names.arg = res$Term[5:1],
#         horiz = TRUE, las = 1, border = FALSE, cex.names = 0.6)
#     abline(v = c(-log10(0.05)), lty = 2)
#     abline(v = 0, lty = 1)
# }





```

# Analysis


## Integrated space{.tabset}


```{r sp_combined, message=FALSE, warning=FALSE, fig.height=8, fig.width=18, results='asis', cache=TRUE}




umap = reducedDim(adata_q, 'X_umap')
p_umap1 = plot_2d(umap, adata_q$section, label_cols=.palette3, label_title='sample') 
p_umap2 = plot_2d(umap, adata_q$leiden_sagenet, label_cols=.palette2, label_title='(integrated) nodule') 
p = p_umap1 + p_umap2

# umap_sagenet = reducedDim(adata_q, 'X_umap_sagenet')
# p_umap_sagenet = plot_2d(umap_sagenet, adata_q$section, label_cols=.palette_all) + ggtitle('SageNet')

# p_umap_sagenet_clus = plot_2d(umap_sagenet, factor(adata_q$leiden_sagenet), label_cols=.palette_all) + ggtitle('SageNet')


# p = p_umap + p_umap_sagenet + p_umap_sagenet_clus + plot_layout(ncol = 3, guides='collect')

print(p)


```


## Spatial maps{.tabset}

```{r sp_all, message=FALSE, warning=FALSE, fig.height=8, fig.width=17, results='asis', cache=TRUE}

leiden_cols = .palette2[as.numeric(as.factor(unique(adata_q$leiden_sagenet)))]
        names(leiden_cols) = unique(adata_q$leiden_sagenet)
leiden_cols = c(leiden_cols, c('normal' = '#999999'))

for(sample_tag in samples_tags){

    

    cat('### ', sample_tag, '{.tabset}\n')


        cols = .palette_all[as.numeric(as.factor(unique(samples[[sample_tag]]$nodule)))]
        names(cols) = unique(samples[[sample_tag]]$nodule)

        

        p_vis_nodule = vis_clus(
            spe = samples[[sample_tag]],
            clustervar = "nodule",
            colors = cols, 
            size=4,
            alpha=0.8,
            sample_id ='initial annotation',
        )  + ggtitle(sample_tag) + theme(legend.position='none') 

        p_vis_leiden = vis_clus(
            spe = samples[[sample_tag]],
            clustervar = "leiden_sagenet",
            colors = leiden_cols, 
            size=4,
            alpha=0.8,
            sample_id ='unified annotation',
        )  + ggtitle('unified annotation') 
       

        p = p_vis_nodule + p_vis_leiden  + plot_layout(ncol = 2, guides='collect')

        print(p)
    
    cat('\n\n')

}



```




## Expression heatmaps of gene modules and nodules{.tabset}

Gene modules come from the Consensus graph. Expression values computed on each slide separately.


### Unified annotation{.tabset}

```{r hm_unified, message=FALSE, warning=FALSE, fig.height=22, fig.width=10, results='asis', cache=TRUE}

for(sample_tag in samples_tags){

    act_leiden = act_all_leiden[[sample_tag]]
    act_leiden = act_leiden[,names(cons_mem)]

    cat('#### ', sample_tag, '{.tabset}\n')
        mtx = t(act_leiden)
        col_cols = leiden_cols[rownames(act_leiden)]
        names(col_cols) = rownames(act_leiden)
        col_annots  = HeatmapAnnotation(
            group                = rownames(act_leiden), 
            col                  = list(group  = col_cols),
            annotation_name_side = 'left', 
            show_legend          = c(group=FALSE)
        )

        row_cols = .palette1[cons_mem]
        names(row_cols) = cons_mem
        # names(row_cols) = comm_dt$community
        row_annots  = rowAnnotation(
            gene_module          = cons_mem, 
            col                  = list(gene_module  = row_cols),
            show_legend          = c(gene_module=FALSE)
        )

        grp_vals  = seq(min(mtx), max(mtx), length.out=9)
        grp_cols  = circlize::colorRamp2(grp_vals, viridis::viridis(9))
        if(dim(as.matrix(mtx))[2] > 1){
            seriate_obj  = seriate(as.matrix(mtx - min(mtx)), method = "BEA_TSP")
            row_order    = get_order(seriate_obj, 1)
            column_order = get_order(seriate_obj, 2)
        }else{
            row_order    = 1:length(mtx)
            column_order = 1
        }

        # do heatmap for these genes
        hm_obj      = Heatmap(
            matrix=mtx, col=grp_cols,
            row_order=row_order, 
            cluster_rows=TRUE, cluster_columns=TRUE,
            row_names_gp=gpar(fontsize = 8), column_names_gp=gpar(fontsize = 10),
            name="Scaled\nGene Expression", 
            row_names_side="left",
            column_names_side="top",
            row_split=cons_mem,
            cluster_column_slices=FALSE,
            top_annotation=col_annots, left_annotation=row_annots
            )
        draw(hm_obj)
    
    
    cat('\n\n')

}
```

### Preliminary annotation{.tabset}

```{r hm_all, message=FALSE, warning=FALSE, fig.height=22, fig.width=10, results='asis', cache=TRUE}

for(sample_tag in samples_tags){

    act_nodule = act_all_nodules[[sample_tag]]
    act_nodule = act_nodule[,names(cons_mem)]

    cat('#### ', sample_tag, '{.tabset}\n')
        mtx = t(act_nodule)
        col_cols = .palette_all[1:nrow(act_nodule)]
        names(col_cols) = rownames(act_nodule)
        col_annots  = HeatmapAnnotation(
            group                = rownames(act_nodule), 
            col                  = list(group  = col_cols),
            annotation_name_side = 'left', 
            show_legend          = c(group=FALSE)
        )

        row_cols = .palette1[cons_mem]
        names(row_cols) = cons_mem
        # names(row_cols) = comm_dt$community
        row_annots  = rowAnnotation(
            gene_module          = cons_mem, 
            col                  = list(gene_module  = row_cols),
            show_legend          = c(gene_module=FALSE)
        )

        grp_vals  = seq(min(mtx), max(mtx), length.out=9)
        grp_cols  = circlize::colorRamp2(grp_vals, viridis::viridis(9))
        if(dim(as.matrix(mtx))[2] > 1){
            seriate_obj  = seriate(as.matrix(mtx - min(mtx)), method = "BEA_TSP")
            row_order    = get_order(seriate_obj, 1)
            column_order = get_order(seriate_obj, 2)
        }else{
            row_order    = 1:length(mtx)
            column_order = 1
        }

        # do heatmap for these genes
        hm_obj      = Heatmap(
            matrix=mtx, col=grp_cols,
            row_order=row_order, 
            cluster_rows=TRUE, cluster_columns=TRUE,
            row_names_gp=gpar(fontsize = 8), column_names_gp=gpar(fontsize = 10),
            name="Scaled\nGene Expression", 
            row_names_side="left",
            column_names_side="top",
            row_split=cons_mem,
            cluster_column_slices=FALSE,
            top_annotation=col_annots, left_annotation=row_annots
            )
        draw(hm_obj)
    
    
    cat('\n\n')

}
```

## Separated gene interaction graphs{.tabset}
```{r ggm_sep_new, fig.height=8, fig.width=8, results='asis', cache=TRUE}
for(sample_tag in samples_tags){

    # lay = layout_nicely(gg[[sample_tag]])

    cat('### ', sample_tag, '{.tabset}\n')

    # cat('#### gene interaction network{.tabset}\n')
    #     cat('##### \n')
    #     graph_col_comm(
    #         graph  = gg[[sample_tag]], 
    #         lay    = lay, 
    #         grp    = .palette1[memberships[[sample_tag]]], 
    #         title  = sprintf('gene interaction network on %s', sample_tag),
    #         labels = names(memberships[[sample_tag]])
    #     )
    #     cat('\n\n')
    # cat('\n\n')

    cat('#### gene activities{.tabset}\n')

    for(l in sort(rownames(act_all_leiden[[sample_tag]]))){
        cat('##### ', l, '\n')
        pp = graph_col_act(cons_graph, act_all_leiden[[sample_tag]][l, names(V(cons_graph))], cons_lay, sprintf('%s: %s', sample_tag, l))
        print(pp)
        cat('\n\n')
    }

    cat('\n\n')
    
    
    cat('\n\n')

}
```



## Separated gene interaction graphs{.tabset}
```{r ggm_sep, fig.height=8, fig.width=8, results='asis', cache=TRUE}
for(sample_tag in samples_tags){

    lay = layout_nicely(gg[[sample_tag]])

    cat('### ', sample_tag, '{.tabset}\n')

    cat('#### gene interaction network{.tabset}\n')
        cat('##### \n')
        graph_col_comm(
            graph  = gg[[sample_tag]], 
            lay    = lay, 
            grp    = .palette1[memberships[[sample_tag]]], 
            title  = sprintf('gene interaction network on %s', sample_tag),
            labels = names(memberships[[sample_tag]])
        )
        cat('\n\n')
    cat('\n\n')

    cat('#### gene activities{.tabset}\n')
    cat('##### jojo\n')
    pp = graph_col_act(cons_graph, act_all_nodules[[sample_tag]][1, names(V(cons_graph))], lay, sprintf('%s: %s', sample_tag, 'test'))
        print(pp)
    cat('\n\n') 

    for(nodule in rownames(act_all_nodules[[sample_tag]])){
        cat('##### ', nodule, '\n')
        pp = graph_col_act(cons_graph, act_all_nodules[[sample_tag]][nodule, names(V(cons_graph))], lay, sprintf('%s: %s', sample_tag, nodule))
        print(pp)
        cat('\n\n')
    }

    cat('\n\n')
    
    
    cat('\n\n')

}
```


# Outputs

```{r save_outputs}
```

```{r session_info, include=TRUE, echo=TRUE, results='markup'}
devtools::session_info()
```
